### 2021-02-24 rad-icfp-2021/intro.tex
improving intro a bit

\begin{itemize}
    \item expressive source language with arrays but restrictive enough for efficient $\Dsynrevsymbol$
    \item method for extending our work and modular method for adding and prove correct new primitives as needed in practice 
    \item new I.R.: UNF
    \item a very simple $\Dsynrevsymbol$ macro
    \item expressive purely functional target language
    \item several generalisation at the end
    \item correctness and complexity theorems
    \item recover efficiency via standard optimisation techniques as opposed to ad-hoc ones
    \item detailed comparison with other work and limitations to our work, some ideas around limitations
\end{itemize}

@@

\MH{forward reference to section 2 where reverse-mode is recalled. maybe same for SSA. }

We introduce a general idea of how to do reverse-mode in a functional setting that is efficient and pure. 
Let's have a look at what we obtain on a few examples.

\begin{example}[First-order term]
On simple first-order terms, the output code looks like SSA code \cite{}.
Let's consider  the term !let w1 = x1 * x2 in let w2 = w1 * x1 in w2! in the context $\Gamma:=\{x1,x2,x3:\RR\}$.\\
After our (inefficient) reverse-mode transformation, we obtain
\begin{center}
    \begin{tabular}{l}
        !let w1,w1'! = !< x1*x2, fun (y1,...,y4) -> (y1+x2*y4, y2+x1*y4, y3)> in!\\
        !let w2,w2'! = !< w1*x1, fun (y1,...,y5) -> w1'!!(y1+w1*y5, y2, y3, y4+x1*y5)> in!\\
        !w_2'(0,0,0,0,1)!
    \end{tabular}
\end{center}
The part $(0,0,0,0,1)$ comes as an equivalent to initialing the tangent variables in the imperative reverse-mode algorithm.

After some general partial evaluation that will be detailed further in the paper, we obtain     
        \begin{center}
            \begin{tabular}{l}
                !let w1 = x1 * x2 in!\\ 
                !let w2 = w1 * x1 in!\\
                !let y1 = 0 in!\\
                !let y2 = 0 in!\\
                !let y3 = 0 in!\\
                !let y4 = 0 in!\\
                !let y5 = 1 in!\\
                !let y1'! != y1+w1*y5 in!\\
                !let y4'! != y4+x1*y5 in!\\
                !(y1'+x2*y4'!!, y2+x1*y4'!!, y3)!
            \end{tabular}
        \end{center}  
This is equivalent to the SSA form of what the imperative reverse-mode differentiation of our initial term would be.
Of course this can be further optimised to give
        \begin{center}
            \begin{tabular}{{c}}
                !let w1 = x1 * x2 in!\\ 
                !let w2 = w1 * x1 in!\\
                !(w1+x2*x1, x1*x1, 0)!
            \end{tabular}
        \end{center}
    \end{example}

>>>>

Our work provides a sort of compilation scheme from a source functional language to a target functional language for reverse-mode differentiation.
To emphasize some key ideas and for presentation purposes, we present our work on a simple yet expressive language in Section~\ref{sec:simplediff}, 
and show some useful generalisation in Section~\ref{sec:generalisation}.

The key ideas of our work can be summarised by saying that reverse-mode differentiation
\begin{itemize}
   \item can be made efficient way more easily on unary operators
   \item can be done efficiently and purely using second-order array operators for second-order array operators (e.g. !map,foldl!) 
   \item can be made efficient via standard functional optimisation techniques
\end{itemize}

Our contribution tries to make these claims precise. Our detailed  contribution consist of
\begin{itemize}
    \item introducing an expressive source language with array operations, yet restrictive enough for efficient differentiation
    \item a method for extending our work and modular method for adding and prove correct new primitives as needed in practice 
    \item a new intermediate representation (I.R.) which we call unary form(UNF), essentially turning every operation as a unary one
    \item a simple reverse-mode transformation on the I.R.
    \item an expressive purely functional target language and efficient transformation from the I.R. to this language
    \item several useful generalisations in Section~\ref{sec:generalisation}. 
    \item correctness of the differentiation \ref{sec:correctness}, and complexity guarantees \ref{sec:complexity}
    \item showing how efficiency  is recovered via standard optimisation techniques
    \item a detailed comparison with other work and some limitations to our work in Section~\ref{sec:conclusion}
\end{itemize}

@@

We introduce a general idea of how to do reverse-mode in a functional setting that is efficient and pure.
We recall some basics of reverse-mode differentiation in Section~\ref{sec:background}.
Let's have a look at what we obtain on a few examples.

\begin{example}[First-order term]
On simple first-order terms, the output code looks like SSA code \cite{}.
Let's consider  the term !let w1 = x1 * x2 in let w2 = w1 * x1 in w2! in the context $\Gamma:=\{x1,x2,x3:\RR\}$.\\
After our (inefficient) reverse-mode transformation, we obtain
\begin{center}
    \begin{tabular}{l}
        !let w1,w1'! = !< x1*x2, fun (y1,...,y4) -> (y1+x2*y4, y2+x1*y4, y3)> in!\\
        !let w2,w2'! = !< w1*x1, fun (y1,...,y5) -> w1'!!(y1+w1*y5, y2, y3, y4+x1*y5)> in!\\
        !w_2'(0,0,0,0,1)!
    \end{tabular}
\end{center}
The part !(0,0,0,0,1)! comes as an equivalent to initialing the tangent variables in the imperative reverse-mode algorithm.

After some general partial evaluation that will be detailed further in the paper, we obtain     
        \begin{center}
            \begin{tabular}{l}
                !let w1 = x1 * x2 in!\\ 
                !let w2 = w1 * x1 in!\\
                !let y1 = 0 in!\\
                !let y2 = 0 in!\\
                !let y3 = 0 in!\\
                !let y4 = 0 in!\\
                !let y5 = 1 in!\\
                !let y1'! != y1+w1*y5 in!\\
                !let y4'! != y4+x1*y5 in!\\
                !(y1'+x2*y4'!!, y2+x1*y4'!!, y3)!
            \end{tabular}
        \end{center}  
For the expert viewer, this is equivalent to the SSA form of what the imperative reverse-mode differentiation of our initial term would be.\\
Of course, this term can be further optimised to give
        \begin{center}
            \begin{tabular}{{c}}
                !let w1 = x1 * x2 in!\\ 
                !let w2 = w1 * x1 in!\\
                !(w1+x2*x1, x1*x1, 0)!
            \end{tabular}
        \end{center}
    \end{example}
